---
title: "MSDS 6372 Project 1: EDA"
author: "Maureen Stolberg"
date: "June 2, 2019"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

## Predicting Wine Preferences based on it's Physicochemical Properties

**
Citation Request:
**

  This dataset is public available for research. The details are described in [Cortez et al., 2009]. 
  Please include this citation if you plan to use this database:

  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems>, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: https://archive.ics.uci.edu/ml/datasets/Wine+Quality

For more information, read [Cortez et al., 2009].

## Data Description

The datasets are publicly available for research purposes and the details are described in [Cortez et al., 2009]. I collected the data from the UCI repository website (https://archive.ics.uci.edu/ml/datasets/Wine+Quality). Dataset is large enough to rely on its result ( with 4898 white and 1599 red entries ) of Vinho Verde (Portugal Wine Company from the northwest region of Portugal) samples are considered:  

.	winequality-red.csv - red wine preference samples
.	winequality-white.csv - white wine preference samples  

These data sets include physicochemical and sensory related data to red and white Vinho Verde wine sample. The data were collected from May/2004 to February/2007 using only protected designation of origin samples that were tested at the official certification entity (CVRVV). The CVRVV is an inter-professional organization with the goal of improving the quality and marketing of Vinho Verde. The data were recorded by a computerized system (iLab), which automatically manages the process of wine sample testing from producer requests to the laboratory and sensory analysis. Each entry denotes a given test (analytical or sensory) and the final database was exported into a single sheet (.csv).

**
Description of attributes:
**  

**Fixed acidity:** most acids involved with wine or fixed or nonvolatile (do not evaporate readily)  

**Volatile acidity:** the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar     taste.  

**Citric acid:** found in small quantities, citric acid can add 'freshness' and flavor to wines  

**Residual sugar:** the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1       gram/liter and wines with greater than 45 grams/liter are considered sweet.  

**Chlorides:** the amount of salt in the wine  

**Free sulfur dioxide:** the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and         bisulfite ion; it prevents microbial growth and the oxidation of wine  

**Total sulfur dioxide:** amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in      wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine  

**Density:** the density of water is close to that of water depending on the percent alcohol and sugar content  

**pH:** describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are         between 3-4 on the pH scale  

**Sulphates:** a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant  

**Alcohol:** the percent alcohol content of the wine  

**Quality:** output variable (based on sensory data, score between 0 and 10)

## Univariate Data Summary Statistics

**Data Summary**  
```{r Question 1, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}


#load my standard toolbox
suppressPackageStartupMessages(library(tidyverse)) 
suppressPackageStartupMessages(library(dataMaid))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(corrplot))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(dlookr))

#Set working directory
setwd("/MSDS6372/winequality")

#load in the red wine data from the UCI ML website
redwine<-read.csv('http://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv',sep=';')

#load in the red wine data from the UCI ML website
whitewine<-read.csv('http://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-white.csv',sep=';')

#Load combined File
totalwine<-read.csv ("C:/MSDS6372/winequality/CombinedFile.csv")

#Select dataset to analyize/comment out other options
#wine.data <- redwine
#wine.data <- whitewine
wine.data <- totalwine

#Convert categorical variables to factors
#2nd factor-wine.category is optional-only run if using totalwine file
wine.data$Quality.Factor <- as.factor(wine.data$quality)
wine.data$wine.category.f <- as.factor(wine.data$wine.category)

# Define original variable names
ORIGINAL <- colnames(wine.data) [1:12]

# Define original + categorical (red/white) variable names
# Optional-valid for totalwine data only.  Otherwise disregard and cancel out command.
explanatory.variables <- colnames(wine.data)[1:11]


#Standardize data
wine.data_scaled <- data.frame(scale(wine.data[,ORIGINAL]))

#Summary Table of data
summary(wine.data)
```

  
  
**Descriptive Statistics (Full Dataset)**

```{r Question 2, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

#Full dataset : Descriptive Statistics- sort by skewness
wine.data %>%
  describe() %>%
  select(variable, skewness, mean, p25, p50, p75) %>% 
  filter(!is.na(skewness)) %>% 
  arrange(desc(abs(skewness)))

```


**Red vs. White Wine: Descriptive Statistics**

```{r Question 3, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

wine.data %>%
group_by(wine.category) %>%
describe() %>%
select(variable,wine.category, skewness, mean, p25, p50, p75) %>% 
filter(!is.na(skewness)) %>% 
arrange(desc(variable))

```

## Normality Review

###Scatterplot Matrix

**Red and White Wine dataset combined**  


```{r Question 4, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
pairs(wine.data)

```


**Red Wine Only**  

```{r Question 4a, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
pairs(redwine, col=redwine$quality)

```

  
  

**White Wine Only**  

```{r Question 4b, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
pairs(whitewine, col=whitewine$quality)

```
  

### Distribution Plots (based on original data)

  Where:
  - Blue line represents the mean
  - Dark magenta line represents the median
  - Dashed line represents the 25th and 75th percentile  
  
```{r Question 5, echo=FALSE}

dens <- lapply(ORIGINAL, FUN=function(var) {
  ggplot(wine.data, aes_string(x=var)) + 
    geom_density(fill='gray') +
    geom_vline(aes(xintercept=mean(wine.data[, var])), color='blue', size=1) +
    geom_vline(aes(xintercept=median(wine.data[, var])), color='darkmagenta', size=1) +
    geom_vline(aes(xintercept=quantile(wine.data[, var], 0.25)), 
               linetype='dashed', size=0.5) + 
    geom_vline(aes(xintercept=quantile(wine.data[, var], 0.75)), 
               linetype='dashed', size=0.5)
})
do.call(grid.arrange, args=c(dens, list(ncol=3)))

```


Review Variable transformation options to determine if data should be transformed prior to model creation.  

Let:  
- origin represents normality based on raw unadjusted data
- Log represents normality based on log transformed data
- sqrt represents normality based on the square root of original data


```{r Question 6, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

#Graph normality prepackage to evaluate transformation options (Origin, log, SqRt)-if needed
plot_normality(wine.data)
```

## Relationship Assessment  

**Total Wine Inventory Review (Red + White)**  

Lets explore how thte potential explanatory variables may be correlated to the response.  

```{r Question 7, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
showBoxplt <- function(df, y, x='Quality.Factor', lower=0, upper=.99) {
  ggplot(df, aes_string(x=x, y=y, fill=x)) + 
    geom_boxplot() + 
    ylim(quantile(df[, y], prob=lower), 
         quantile(df[, y], prob=upper)) +
    theme(legend.position="none")
}
boxp1 <- lapply(explanatory.variables, FUN=function(var) showBoxplt(wine.data, var))
do.call(grid.arrange, args=c(boxp1, list(ncol=3)))

```


**Red Vs. White Wine Review**  

```{r Question 8, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
boxp3 <- lapply(explanatory.variables, 
                FUN=function(var) showBoxplt(wine.data, y=var, x='wine.category'))
do.call(grid.arrange, args=c(boxp3, list(ncol=3)))

```




### Correlation Matrix  

```{r Question 9, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Plot correlation matrix to determine is collinearity is present
wine.data %>%
plot_correlate(explanatory.variables)
```
